---
title: "The Effect of Transmission Type on MPG"
author: "Paul Martinez"
date: "04/10/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5)
options(width=120)
```

## Executive Summary

The following analysis examines the relationship between transmission type and 
MPG. This analysis answers the following two questions:

* “Is an automatic or manual transmission better for MPG”
* "Quantify the MPG difference between automatic and manual transmissions"


## Data Upload and Cleaning

I convert some variables to factor to facilitate with the analysis. 

```{r}
library(ggplot2)
library(dplyr)
data(mtcars)

mtcars2<-mtcars


mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c("automatic", "manual")
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$vs <- as.factor(mtcars$vs)
levels(mtcars$vs) <- c("V", "S")

```

## Exploratory data analyses 

Next, I explore the data.
```{r}

head(mtcars)
```


The following scatterplot examines the relationship between transmission type and MPG. 
From these visualization we can see that there is a significant difference between automatic
and manual transmissions. It appears manual transmissions are slightly more efficient when 
it comes to MPG. We will conduct further testing. 

```{r}
# Figure 1
ggplot(data = mtcars,   # set data
       mapping = aes(     # map aesthetics to column values
               x = mpg,           # map x-axis to age            
               y = am,         # map y-axis to weight
               color = am)
)+     # map color to age
        geom_point() 

```

Prior to creating a model, I wanted to explore if other variables also had a similar relationship with MPG.
After conducting the Pearson correlation test below, we can see that the weight(wt) of the car and how
many cylinders also had an extremely significant relationship between how much MPG is used and those variables. 

```{r}
# Correlation Test
cors <- cor(mtcars2$mpg, mtcars2)
orderedCors <- cors[,order(-abs(cors[1,]))]
orderedCors

```


## Model selection 


To start I test the question “Is an automatic or manual transmission better for MPG” using a linear regression model.
In this model I only include MPG as my dependent variable and transmission type as my independent variable. 
The results show that manual transmission in comparison to automatic yields a 7.24 positive MPG. This result was statistically
significant at a p-value less than 0.000.

```{r}
# MODEL 1- TRANSMISSION TYPE

model1 <- lm(mpg ~ am, mtcars)
summary(model1)
res <- resid(model1)

#Residual Plot for Model 1
plot(fitted(model1), res)

```

Next, while Model 1 showed promising results I wanted to explore what would happen if we add additional factors.
Holding constant all other variables in the dataset, transmission type was not statistically signficant in model 2. In fact, 
the only two variables that showed to have a statistically significant value at p>0.05 were the Weight of the car and gross horsepower, 
while holding all other variables constant.

```{r}
# MODEL 2-All VARIABLES
model2 <- lm(mpg ~ ., mtcars)
summary(model2)
res2 <- resid(model2)

#Residual Plot for Model 2
plot(fitted(model2), res2)
```


## Conclusion

In conclusion, if we only considered Model 1 the answer to the question would be absolutely transmission type matter, 
when considering MPG. However, if we account for other factors, we can see that it is not a statistically significant factor
in comparison to other variables like the Weight of the car or the horsepower. 
